Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-03-29 draft.tex
Added two reworked figures

Morphometric variables are derived to allow for statistical comparison of the tissue, which incorporate mostly positions, diameters, volumes and relative distances and alignment to other structures.
% 
% - Data Format\\
% - Missing Data\\
% - Registration of Image Data (VMV'12)\\
% - Extraction of Image Parameter
\subsection{Analysis workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/InteractionLoop}
 % TODO: Recreate Image: Also include Cramér´s V which is part of the visualization
 \caption{Detailed Version of the \emph{IVA Loop}. Usually starting with a selection of Feature of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types. The data are visualized in the range and domain space, which can then be brushed, yielding new groups, which can be investigated using further features. Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
 \label{fig:InteractionLoop}
\end{figure}
Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, -visualization and -brushing.
%
An hypothesis-driven analysis usually starts with the selection of features.
%
A feature selection can also be derived from a shape-based clustering, which creates shape groups.
%
\emph{Hypothesis-generation} with focus on image data starts with a shape-based clustering.
%
The feature is mapped using a automatically chosen visualization appropriate for its data type (described in detail in the following section).
%
The visualization techniques have to combine both image-and non-image data in order to set domain and range data in relation to each other.
%
Subsequently the visualization can be either brushed or new variables can be added to the analysis.
%
Brushing methods are subdivided using the previously described \emph{IVA} patterns.
%
Brushed regions are treated like features as they divide the subject space just like categorical features. %and lead therefore to a new feature selection.
%
Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight variables interacting with selected features.
%
%The starting point of \emph{IVA Loop} is determined whether the analysis is hypothesis-driven or exploratory.
%
%An analysis with the aim of hypothesis generation of image-based data starts with a clustering of the structure 
%
A sample workflow using interaction- and visualization techniques described in the next section can be seen in Figure~\ref{fig:visualization}.

\subsection{Interaction- and Visualization Techniques} \label{Interaction- and Visualization Techniques}
%-------------------------------- Figure --------------------------------
% TODO: VISUALIZATION Figure sollte umgebaut werden, sodass sie die Schritte bei der Analyse hervorhebt
\begin{figure*}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 % \includegraphics[width=3.5in]{figures/visualization}
 %\includegraphics[width=1\textwidth]{figures/visualization}
 \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
 \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) Sidebar containing all features as well as the groups defined in the analysis process; (b) Canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) Interactive Pivot Table showing the exact numbers for each displayed variable combination. The data displayed is used to analyze the lumbar spine. Features can be added freely on the canvas via drag and drop. Dropping the \emph{age} parameter on the already plotted \emph{body size} container creates a mosaic plot combining both features (right).
 }
 \label{fig:visualization}
\end{figure*}
%-------------------------------- /Figure -------------------------------
%We employ the fourth and highest level which includes next to brushing and linking, advanced brushing, extraction of attributes the development of visualizations custom tailored to the data sets.
%
%\emph{IVA} levels define different levels interaction. 
%
The suitability of an interaction- and visualization technique for epidemiological data depends on its ability to intuitively compare multiple data features at once while highlighting new associations.
%
%Since epidemiologists come from different expertise, their different background has to be taken into account.
%
The chances of a method to be adapted by epidemiologists increases highly if their domain language is used and the methods reflect routines they are used to.
%
Visual evaluation of data are therefore as important as methods allowing for numerical data analysis.
%
In the following sections we present the different parts of our proposed \emph{IVA} system for image-based cohort study analysis.
%Chosing or creating well suited visualization- anassociationon techniques for epidemiological data
%The employed techniques need to respect the epidemiological requirements.

@@

\subsection{Implementation}
\begin{figure}[htb]
 \centering
 \label{fig:technologies}
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/technologies}
 \caption{\emph{Image is not final}. The front-end solution (left) uses state of the art web technologies such as HTML5/CSS3, WebGL and SVG to display the data.
 % 
 The NodeJS based back-end (right) stores all image- and non-image data and transfers it to connected clients.
 %
 All computation heavy operations like calculation of mean-shapes or -distances as well as statistical processing is done by the server to keep hardware requirements of client systems low. 
 %
 Client-Server communication is accomplished via the WebSocket protocol.
 }
\end{figure}
%
To provide a fast communication loop between method development and expert input, we decided to rely on modern web technologies, which benefits from various advantages:
\begin{itemize}
	\item No additional software needs to be installed, most people use decent state-of-the-art web browser, even on mobile devices.
	\item The client-server structure allows for employing heavy computation on a server machine and transfer results to the client.
	\item Since image data for several thousand subjects requires hundreds of gigabytes disk space it can remain safely on the server and elements can be transferred on demand.
	%
	High confidentiality standards of the data can be met by restricting access via a account system
	\item Recent developments in WebGL applications running in browsers with near native performance push the development into to the web resulting in many open source libraries, which are well documented, rich in examples and driven by active communities. We use WebGL for rendering shape information.
	\item Web technologies are free.
\end{itemize}
These advantages do not come without drawbacks.
%
%Many methods which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics have build in, need to be written from scratch in order to fit in the context.
Due to missing ports of sophisticated specialized libraries/languages, such as the Visualization Toolkit (VTK) or R for statistics, many standard methods need to be written from scratch for the web.
%Many methods, which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics, have build in, need to be written from scratch in order to fit in the context.
%

>>>>

Morphometric variables are derived to allow for statistical comparison of the tissue, which incorporate mostly positions, diameters, volumes and relative distances and alignment to other structures.
% 
% - Data Format\\
% - Missing Data\\
% - Registration of Image Data (VMV'12)\\
% - Extraction of Image Parameter
\subsection{Analysis workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/InteractionLoop}
 % TODO: Recreate Image: Also include Cramér´s V which is part of the visualization
 \caption{Detailed Version of the \emph{IVA Loop}. Usually starting with a selection of Feature of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types. The data are visualized in the range and domain space, which can then be brushed, yielding new groups, which can be investigated using further features. Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
 \label{fig:InteractionLoop}
\end{figure}
%-------------------------------- Figure --------------------------------
\begin{figure*}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 % \includegraphics[width=3.5in]{figures/visualization}
 %\includegraphics[width=1\textwidth]{figures/visualization}
 \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
 \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) Sidebar containing all features as well as the groups defined in the analysis process; (b) Canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) Interactive Pivot Table showing the exact numbers for each displayed variable combination. The data displayed is used to analyze the lumbar spine. Features can be added freely on the canvas via drag and drop. Dropping the \emph{age} parameter on the already plotted \emph{body size} container creates a mosaic plot combining both features (right).
 }
 \label{fig:visualization}
\end{figure*}
%-------------------------------- /Figure -------------------------------
Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, -visualization and -brushing.
%
An hypothesis-driven analysis usually starts with the selection of features.
%
A feature selection can also be derived from a shape-based clustering, which creates shape groups.
%
\emph{Hypothesis-generation} with focus on image data starts with a shape-based clustering.
%
The feature is mapped using a automatically chosen visualization appropriate for its data type (described in detail in the following section).
%
The visualization techniques have to combine both image-and non-image data in order to set domain and range data in relation to each other.
%
Subsequently the visualization can be either brushed or new variables can be added to the analysis.
%
Brushing methods are subdivided using the previously described \emph{IVA} patterns.
%
Brushed regions are treated like features as they divide the subject space just like categorical features. %and lead therefore to a new feature selection.
%
Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight variables interacting with selected features.
%
%The starting point of \emph{IVA Loop} is determined whether the analysis is hypothesis-driven or exploratory.
%
%An analysis with the aim of hypothesis generation of image-based data starts with a clustering of the structure 
%
A sample workflow using interaction- and visualization techniques described in the next section can be seen in Figure~\ref{fig:visualization}.

\subsection{Interaction- and Visualization Techniques} \label{Interaction- and Visualization Techniques}
%We employ the fourth and highest level which includes next to brushing and linking, advanced brushing, extraction of attributes the development of visualizations custom tailored to the data sets.
%
%\emph{IVA} levels define different levels interaction. 
%
The suitability of an interaction- and visualization technique for epidemiological data depends on its ability to intuitively compare multiple data features at once while highlighting new associations.
%
%Since epidemiologists come from different expertise, their different background has to be taken into account.
%
The chances of a method to be adapted by epidemiologists increases highly if their domain language is used and the methods reflect routines they are used to.
%
Visual evaluation of data are therefore as important as methods allowing for numerical data analysis.
%
In the following sections we present the different parts of our proposed \emph{IVA} system for image-based cohort study analysis.
%Chosing or creating well suited visualization- anassociationon techniques for epidemiological data
%The employed techniques need to respect the epidemiological requirements.

@@

\subsection{Implementation}
\begin{figure}[htb]
 \centering
 \label{fig:technologies}
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/technologies}
 \caption{The front-end solution (left) uses state of the art web technologies such as HTML5/CSS3, WebGL and SVG to display the data.
 % 
 The NodeJS based back-end (right) stores all image- and non-image data and transfers it to connected clients.
 %
 All computation heavy operations like calculation of mean-shapes or -distances as well as statistical processing is done by the server to keep hardware requirements of client systems low. 
 %
 Client-Server communication is accomplished via the Websocket protocol.
 }
\end{figure}
%
To provide a fast communication loop between method development and expert input, we decided to rely on modern web technologies, which benefits from various advantages:
\begin{itemize}
	\item No additional software needs to be installed, most people use decent state-of-the-art web browser, even on mobile devices.
	\item The client-server structure allows for employing heavy computation on a server machine and transfer results to the client.
	\item Since image data for several thousand subjects requires hundreds of gigabytes disk space it can remain safely on the server and elements can be transferred on demand.
	%
	High confidentiality standards of the data can be met by restricting access via a account system
	\item Recent developments in WebGL applications running in browsers with near native performance push the development into to the web resulting in many open source libraries, which are well documented, rich in examples and driven by active communities. We use WebGL for rendering shape information.
	\item Web technologies are free.
\end{itemize}
These advantages do not come without drawbacks.
%
%Many methods which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics have build in, need to be written from scratch in order to fit in the context.
Due to missing ports of sophisticated specialized libraries/languages, such as the Visualization Toolkit (VTK) or R for statistics, many standard methods need to be written from scratch for the web.
%Many methods, which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics, have build in, need to be written from scratch in order to fit in the context.
%
